\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \footnote{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-10-06 reviewer2_jact_response.txt

Reviewer 1:

Reviewer 2
I believe the paper contains very good and interesting results, but it needs a revision before being published.

Pros: I really like the main idea of the paper: using the machinery of moves that allows one to efficiently recompute the R=DV decomposition after one cyclic permutation in the input filtration, the authors find a heuristic to decompose an arbitrary permutation of the filtration into a sequence of moves of provably minimal length and small cost (this is the heuristic part). The authors show experimentally that this idea works. The paper is as self-contained as possible and the authors work out some concrete examples which are helpful to the reader.

Cons: The detailed list of the remarks follows. Sorry, it’s not ordered by importance, so there are trivial things like grammar/typos and more serious issues.
 In general, I feel that the paper is a bit too verbose - it can be made shorter without losing virtually any content.
The most important issues that are listed below are: the comment about ‘MoveLeft’ and some missing (or hard to find) details about the algorithm (GreedySchedule in Alg. 1).
Also, I would appreciate if the authors formulated  exactly what is the implementation they used in their experiments (e.g., did they really use treaps?).

----

Point by point responses:

Page 1, abstract: I don’t understand the $O(m \log \log m)$ estimate here. According to Thm. 1, this is the time it takes just to find $d$, the length of the optimal schedule. Computing the schedule itself costs $O(dm \log m)$ or $O(m \log m)$ per update. Shouldn’t this be in the abstract? Also, why ‘simulation of persistence’ (this expression will appear below, too)?

* - Yes, the $O(m \log \log m)$ complexity refers to the complexity of finding the length of the schedule and the LIS (only). We leave out the complexity of constructing the schedule itself in the abstract as we envision there are situations where the algorithm used to actually construct the schedule may vary---depending on the particular data structure of choice supporting the operations outlined in 3.3 (e.g. implicit treaps, segment tree, vEB tree), the deterministic and expected time complexity may change.

Page 4, 'if the relations are available a-priori' – do the authors mean ‘edges’?

- Yes.

Page 7, ‘we summarize our main results (Theorem 1)’. The main claim in theorem 1 is that one can efficiently compute only the length of the optimal schedule. Will the authors consider reorganizing?

- see above.

Page 8: Although $d$ can be  $O(m)$, we provide evidence that $d ~ m - sqrt(m)$. Well, $m - \sqrt(m)  = O(m)$, why ‘although’? Should it be ‘Although d can be O(m) and we provide evidence that d ~ m - sqrt(m), d is much smaller in practice’?

- Yes, as the previous reviewers also pointed out, we aware that $m - \sqrt(m)  = O(m)$, which is why we omitted the big-O in the notation $d ~ m - sqrt(m)$. To remedy this, we amended the statement to clarify.

Page 8: ‘an simplexwise’ -> a simplexwise

- fixed.

Page 9: I am more used to the ‘positive/negative simplex’ terminology than ‘creator/destroyer’.
This is a matter of personal preference, of course.

-

Page 10: ‘Despite this high algorithmic complexity, the number of column operations has been observed to be super-linear in practice’ – this has to be rephrased, because there is no contradiction (order of m^2 column operations is super-linear, too).

- TODO: cite ripser

Sec. 2.3, Moves. Sorry, it might be my fault but I had difficulty understanding the procedure from the description. I believe having the pseudocode right here will be helpful (after I read it in the appendix, I was happy with the algorithm). Especially since the algorithm is needed in the proof of Prop. 2, too.

- We previously moved much of pseudocode to the appendix in order to streamline the presentation of the concepts, as the pseudocode for both algorithms takes up an entire page (and the pseudocode for MoveRight is also available in Busaryev's paper). We have moved them back. 
(todo: fix?)

Page 16, Alg. 1. 
a) The authors say ‘Algorithm 1 may be easily modified to be completely online, keeping at most two filtrations and one decomposition in memory’. I agree, but I believe it would be equally easy and more natural to state it in this form immediately. The change is in modifying the meaning of \rho_i: while now it is the permutation taking f_1 to f_i, it should be replaced with the permutation taking f_{i-1} to f_i.

- The algorithm is now presented in the much simpler "online form", which only requires as input a pair of filtrations and a single D = RV decomposition. The original batch-processing presentation of pseudocode reflected more closely the implementation we used (we found it more efficient to compute a batch of schedules, concatenate them, and then give them to the moving algorithm to process sequentially, saving the diagrams between each schedule execution).

b) Is it correct that the authors use both MoveRight and MoveLeft in their implementation? I believe it is necessary, and Alg. 1 reflects that; however, it raises some questions about the sequel. I’ll get back to this point.

- Yes!

c) Line 6 calls the function GreedySchedule. There is no such algorithm in the paper. There is Alg. 5 in the appendix, LCS-sort, but there is no greedy heuristic in it. Also, it computes the LIS internally, while GreedySchedule takes the longest increasing sequence as a second argument. Please fix this.

- The GreedySchedule algorithm wasn't included in part because it corresponds to simply modifying line 6 in Alg. 5 (choosing the next 'symbol'/simplex to move) with equation (22).
(todo: mention this or actually build out the pseudocode for 3.4.2)   (TODO... make it explicit)

Page 17, the paragraph after Def. 1: ‘the former can be that can be solved optimally’ - this sentence should be fixed.

- Fixed.

Page 19: ‘small changes … affects’ -> affect.

- Fixed

Note that $m_{i,j}$ is defined here and used in Alg. 5 at the very end of the paper without explanation.

- The usage in Alg. 5 was intentional as it defined in equations (10) and (11); a reference to those equations has been added in Alg. 5.

Page 20: ‘allows use to exploit’ -> allows us to exploit

- Fixed.

Page 21: Do I understand correctly that this is what is going on in the proof of Corollary 1: $E(m - |LCS(p, q)|) = m - 2 \sqrt(m) - cm^{⅙} + o(m^{⅙})$ by the known result of Baik for $E|LIS(p)|$, therefore it is less than $m - \sqrt(m)$ ? So any function which is asymptotically less than $ 2 \sqrt(m) - cm^{⅙} + o(m^{⅙} $ would do? Another question : Corollary 1 speaks about two random filtrations, which the authors interpret as ‘two permutations chosen uniformly at random’ (explaining why this is okay in light of Boissonat’s result). But the longest increasing sequence result by Baik is for a single permutation drawn uniformly at random. Is it obvious that, if p and q are drawn from S_m uniformly at random, then $p \circ q^{-1}$ (the permutation we should consider for LIS) is also uniformly distributed? Well, for symmetry reasons it seems to be true and obvious, but, given the level of details in the other parts of the paper, the authors may want to include this as well.

- We agree that $p \circ q^{-1}$ is uniform, and we agree the proof strategy for Corollary 1 outlined by the reviewer would match the level of detail of the rest of the paper. That said, in our struggle to keep the length of the paper manageable we chose to present the proof in a succinct and informal way.

Page 22, ‘It is clear from Corollary 1’ -  wrong reference. Probably, from the proof of proposition 3 (Prop. 3 itself does not say anything about LIS).

- Yes, this used to be its own corollary. The proof of prop 3 is indeed the reduction we're referring too. We've amended the statement to explicitly point to the proof.

Let T denote a ordered set-like -> an ordered set-like

- Fixed.

Page 27: equation (22) may solved  -> equation (22) can be solved

- Fixed.

General comment on the greedy heuristic: does most of the reasoning in section 3.4 apply to the case of MoveRight only? The starting point is the term \sum |I_k| + |J_k|, which only applies to moving simplex i to simplex j if i < j. What about the MoveLeft? Can it be analyzed theoretically? Does it occur frequently in the authors’ experiments? Can the freedom that the authors have in choosing the schedule of the optimal length be used to minimize the number of left moves and, if I want to implement this approach, should I do that or would it only harm me?

- We actually included a discussion and small analysis of MoveLeft, including some empirical comparisons, in the original draft of the page. To answer your questions successively:

* "does most of the reasoning ... apply to the case of MoveRight only?"

	- Yes, due to theoretical limitations, the statement only holds for MoveRight.

* "What about the MoveLeft? Can it be analyzed theoretically?"

	- We have a small discussion of MoveLeft in the appendix. We left out 
	Our internal

Sec. 4.1: I find the phrase ‘simulate persistence’ unusual. Does it just mean ‘compute persistent homology’? As for the second test (page 28): was the Algorithm 1 used verbatim, or was the scheduling performed to move from one snapshot to the next one?

Page 32, 4.3.1:  Is  \kappa from [10] the same as defined in Theorem 1? I suspect not, [10] says it’s coarseness. Please clarify.

Page 34: ‘they a convenient basis for D may be obtained’ - is this supposed to be ‘they prove that a convenient basis …’ ?
voronoi  -> Voronoi
7-spheres volume -> 7-sphere’s volume
In order reveal -> In order to reveal

Page 35: ‘Projecting the image data onto the first two basis vectors of leads’ - basis vectors of what?

Page 36: ‘the number of permutations applied throughout the encountered along the traversal of the dual graph’ - please fix this sentence.
vineyards is, moves requires -> either ‘vineyards/moves algorithm/approach is/requires’, or ‘vineyards are/moves require’. Or make vineyards and moves boldface or italics, then they are visually perceived as names.
that naively computing -> than naively computing

Page 37: if the filtrations in the parameterization family is nearby -> are nearby

Page 43 , A1.1.1 'After setting V is set' -> After setting V
'Since there exists complexes' -> Since there exists a complex / there exist complexes

Page 45. We recall an important claim .. - What is the claim? Seems to be missing, it should say something about positive/negative simplices.

Page 48. Well, in A.2.1 the authors mention \kappa as coarseness parameter! Also, the naming clash between the structure T used to compute the sorting and the barcode template T is unfortunate.

kendall -> Kendall

-----

readability - We switched the notation from symbol to permutations / indices in section 3.4, and we've reworked the pseudocode ; symbols 
on the overhead of moves - we agree that the figures may seem misleading; future work, output / geometry sensitive . Talk about balance the fraction  of kendall distances
assessment - talk abotu vineyards and non-trivial implementation details; it woudl portray vineyards inaccuractely

Figure 1 - fixed with upper level set filtration or equiv lower filtatratin with 1 - intensity

### 2023-10-06 move_schedules.tex
Addressing most of reviewer 2's comments

%\documentclass{article} % for arxiv only
%\usepackage{arxiv} % for arxiv only 
%\pagestyle{plain} % arxiv only

\documentclass[sn-mathphys]{sn-jnl}

@@

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract} % arxiv only 
\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computationally, simulating persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations that requires only $O(m \log \log m)$ time and $O(m)$ space to construct. 
By reduction to a particular longest common subsequence problem, we show the storage needed to employ this strategy is actually sublinear in expectation. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations, is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
} % springer only
%\end{abstract} % arxiv only

@@

If  zeroth homology   is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the relations are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

 \subsection{A Motivating Example}\label{sec:motivation} 
We  will use a simple experiment to illustrate why the vineyards algorithm does not always yield an efficient strategy for time-varying settings. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, and its associated sublevel-set filtrations, as shown in Figure~\ref{fig:vidExample}.

@@

We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
Since it is unknown a priori when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions based on the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is not guaranteed to capture all homological changes in persistence that occur in simulating a continuous 1-parameter family of diagrams.

@@

We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$  can be $O(m)$ for pathological inputs, we provide evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations   (Corollary~\ref{cor:expectation}), and we give empirical results suggesting $d$  can be much smaller for time-varying filtrations. 

@@

\section{Background}\label{sec:background} 
%\subsection{Notation}\label{sec:notation} 
Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. 
Such a family is called a \emph{filtration},
which is deemed \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Equivalently, a filtration can be also defined as a pair $(K, f)$ where $K$ is a simplicial complex and $f : K \to I$ is a \emph{filter function}  satisfying $f(\tau) \leq f(\sigma)$ in $I$, whenever $\tau \subseteq \sigma$ in  $  K$. In this setting, $K_i = \{ \, \sigma \in K :  f(\sigma) \leq i \, \}$. Here, we consider two index sets: $[m] = \{ 1, \dots, m\}$ and $\mathbb{R}$. 
Without loss of generality, we exclusively consider simplexwise filtrations, but for brevity-sake refer to them simply as filtrations.

@@

 Given a filtration $(K, f)$ with $m$ simplices,  the main output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagrams are encoded in $R$ and the generating cycles in the columns of $V$.
 To begin the reduction, one first   assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. Setting $V = I$ and $R = D$, one proceeds by performing elementary left-to-right column operations on $V$ and $R$ until the following invariants are satisfied:
\textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $(K,f)$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 	\end{enumerate} 
 \vspace*{1em}
 where $\mathrm{low}_R(i)$ denotes the largest row index of a non-zero entry in column $i$ of $R$. 
We call the decomposition satisfying these three invariants \emph{valid}. Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}. The persistence diagrams of the corresponding filtration  can be determined from the lowest entries in $R$, once it has been reduced.

@@

Despite this high algorithmic complexity, the number of column operations has been observed to be super-linear in practice, due in part to the high sparsity and structure of $D$. 
Moreover, many variations and optimizations to Algorithm~\ref{alg:reduce} have been proposed over the past decade, see~\cite{bauer2017phat, bauer2021ripser, chen2011persistent} for an overview. 

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, detecting homological critical events along the way.
As homological critical events only occur when the filtration order changes, detecting all such events may be reduced to computing valid decompositions at   time points interleaving all changes in the filtration order. 
For simplexwise filtrations, these changes manifest as transpositions of adjacent simplices, and thus any fixed set of rules that maintains a valid $R = D V$ decomposition under adjacent column transpositions, is sufficient to simulate persistence dynamically.

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
Formally, since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of vineyards is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the $V$ and $R$ matrices tend to be quite sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. As a rule of thumb, most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

For completeness, we recapitulate the motivation of the moves algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition, where $R = [r_1, r_2, \dots, r_m]$ and $V = [v_1, v_2, \dots, v_m]$ denote the columns of $R$ and $V$, respectively. By definition, if $r_j = D v_j = 0$, then the inclusion $K_{j-1} \hookrightarrow K_{j}$ introduced a new cycle whose generator is given by:
%the non-zero coefficients $(\alpha_i^{(j)}, \alpha_{i+1}^{(j)}, \dots, \alpha_j^{(j)})$ via the combination: 
\begin{equation}\label{eq:coefficients}
	r_j = D v_j = \alpha_j^{(1)} \cdot \partial(\sigma_{1}) + \alpha_{j}^{(2)} \cdot \partial(\sigma_{2}) + \dots + \alpha_{j}^{(j)} \cdot \partial(\sigma_{j}) = 0
\end{equation}
Now, consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, all of the non-zero coefficients $\alpha_k^{(i)}$ contributing to $Dv_k$ for $k \in [i+1, j]$ must be set to zero, as otherwise $\partial(\sigma_i)$ may participate in boundary chains earlier than its own birth in the permuted filtration. 
Setting these coefficients to zero amounts to setting row entries $v_k(i) = 0$ for all $k \in [i+1,j]$---these cancellations manifest as column operations $(\ast)$ used to ensure $P V(\ast)$ is upper-triangular, thus maintaining invariant I2.
Notice that these operations may yield an unreduced $R' = P R(\ast) P^T$, breaking invariant I3.
We could reduce $R'$ with an additional $k$ operations, with the possibility that $k \sim O(\lvert i - j\rvert^2)$, but then this is no more efficient than re-running the reduction algorithm on columns $[i,j]$ in $R$.

@@

\vspace{-3.0em}

@@

Perhaps surprisingly, small changes to set of allowable edit operations $\Sigma$ dramatically affects both the size of $d_\Sigma(p,q)$ and the difficulty of obtaining a minimal sorting. 
For example, while sorting by transpositions and reversals is NP-hard and sorting by prefix  transpositions is unknown, there are polynomial time algorithms for sorting by block interchanges, exchanges, and prefix exchanges~\cite{labarre2013lower}. Sorting by adjacent transpositions can be achieved in many ways: any sorting algorithm that exchanges two adjacent elements during its execution (e.g. bubble sort, insertion sort) yields a sorting of size $K_\tau(p, q)$.

@@

\setlength\arraycolsep{1.8pt}

@@

(\enspace i \enspace j \enspace j\text{-}1 \enspace \dots \enspace i\text{+}2 \enspace i\text{+}1 \enspace )

@@

However, if the pair of filtrations $(K, f), (K, f')$ come from the same underlying complex $K$, then their filter function $f, f'$ may both be thought of a permutations in $S_m$---the corresponding edit distance $d$ then reduces to the \emph{permutation edit distance} problem.

@@

Establishing a connection between the permutation edit distance and move scheduling allows use to exploit the combinatorial structure that comes from the developed theory on both LCS's and LIS's, which are both well-studied objects. We record a single corollary to demonstrate this fact. 

@@

\noindent Note the quantity from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ between pairs of uniformly sampled permutations, as opposed to uniformly sampled filtrations, which have more structure. 

@@

The Spearman distance shares any similarities with $K_\tau$: it is a metric on $S_m$ that is invariant under consistent relabeling. 

@@

    \includegraphics[width=0.80\textwidth]{crossings.png}

>>>>

%\documentclass{article} % for arxiv only
%\usepackage{arxiv} % for arxiv only 
%\pagestyle{plain} % arxiv only 
%%\usepackage[margin=1in]{geometry} % arxiv only

\PassOptionsToPackage{noend}{algorithmic}
\PassOptionsToPackage{noend}{algpseudocode}
\documentclass[sn-mathphys]{sn-jnl}

@@

%\usepackage[noend]{algorithmic}

@@

% REQUIRED - "Abstracts must be able to stand alone and so cannot contain citations to the paper's references, equations, etc. .... not exceeding 250 words"
%\begin{abstract} % arxiv only 
\abstract{ % springer only
Matrix reduction is the standard procedure for computing the persistent homology of a filtered simplicial complex with $m$ simplices. Its output is a particular decomposition of the total boundary matrix, from which the persistence diagrams and generating cycles are derived. 
	Persistence diagrams are known to vary continuously with respect to their input, motivating the  study of their computation for time-varying filtered complexes. Computing persistence dynamically can be reduced to maintaining a valid decomposition under adjacent transpositions in the filtration order. 
	Since there are $O(m^2)$ such transpositions, this maintenance procedure exhibits limited scalability and often is too fine for many applications. 
We propose a coarser strategy for maintaining the decomposition over a 1-parameter family of filtrations. By reduction to a particular longest common subsequence problem, we show that the minimal number $d$ of updates to the decomposition can be found in $O(m \log \log m)$ time and $O(m)$ space, and that the corresponding sequence of permutations---which we call a \emph{schedule}---can be constructed in $O(d m \log m)$ time.  
We also show that, in expectation, the storage needed to employ this strategy is actually sublinear in $m$. 
Exploiting this connection, we show experimentally that the decrease in operations to compute diagrams across a family of filtrations, is proportional to the difference between the expected quadratic number of states and the proposed sublinear coarsening.
Applications to video data, dynamic metric space data, and multi-parameter persistence are also presented.
} % springer only
%\end{abstract} % arxiv only

@@

If  zeroth homology  is the only dimension of interest, computing and updating both the persistence and rank information  is greatly simplified. For example, if the edge are available a-priori, obtaining a tree representation fully characterizing the connectivity of the underlying space (also known as the \emph{incremental connectivity} problem) takes just $O(\alpha(n) n)$ time using the disjoint-set data structure, where $\alpha(n)$ is the extremely slow-growing inverse Ackermann function. 
Adapting this approach to the time-varying setting, Oesterling et al.~\cite{oesterling2015computing} give an algorithm that maintains a \emph{merge tree} with $e$ edges in $O(e)$ time per-update.
If only Betti numbers are needed, the zeroth-dimension problem reduces even further to the \emph{dynamic connectivity problem}, which can be be efficiently solved in amortized $O(\log n)$ query and update times using either Link-cut trees or multi-level Euler tour trees~\cite{kapron2013dynamic}.

\subsection{A Motivating Example}\label{sec:motivation} 
To motivate this effort, we begin with an illustrative example of why the vineyards algorithm does not always yield an efficient strategy for time-varying settings. 
Consider a series of grayscale images (i.e. a video) depicting a fixed-width annulus expanding about the center of a $9 \times 9$ grid, and its associated sublevel-set filtrations, as shown in Figure~\ref{fig:vidExample}.

@@

We use this data set as a baseline for comparing vineyards and the standard reduction algorithm $\texttt{pHcol}$ (Algorithm \ref{alg:reduce}). Suppose a practitioner wanted to know the major homological changes a time-varying filtration encounters over time.
Since it is unknown a priori when the persistent pairing function changes, one solution is to do $n$ independent persistence computations at $n$ evenly spaced points in the time domain.  An alternative approach is to construct a homotopy between a pair of filtrations $(K, f)$, $(K,f')$ and then decompose this homotopy into adjacent transpositions based on the filtration order---the vineyards approach. 
 We refer to the former as the \emph{discrete setting}, which is often used in practice, and the  latter as the \emph{continuous setting}. Note that though the discrete setting is often more practical, it is not guaranteed to capture all homological changes in persistence that occur in the continuous 1-parameter family of diagrams.

@@

We also show that the information needed to construct any $\mathcal{S}$ with optimal size can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. We provide evidence  that $d \sim m - \sqrt{m}$ in expectation for random filtrations (Corollary~\ref{cor:expectation}). Although this implies $d$ can be $O(m)$ for pathological inputs, we give empirical results suggesting $d$  can be much smaller in practice. 

@@

\section{Background}\label{sec:background} 
%\subsection{Notation}\label{sec:notation} 
Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. 
Such a family is called a \emph{filtration},
which is deemed \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into a simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Equivalently, a filtration can be also defined as a pair $(K, f)$ where $K$ is a simplicial complex and $f : K \to I$ is a \emph{filter function}  satisfying $f(\tau) \leq f(\sigma)$ in $I$, whenever $\tau \subseteq \sigma$ in  $  K$. In this setting, $K_i = \{ \, \sigma \in K :  f(\sigma) \leq i \, \}$. Here, we consider two index sets: $[m] = \{ 1, \dots, m\}$ and $\mathbb{R}$. 
Without loss of generality, we exclusively consider simplexwise filtrations, but for brevity-sake refer to them simply as filtrations.

@@

 Given a filtration $(K, f)$ with $m$ simplices,  the output of the reduction algorithm is a matrix decomposition $R = D V$, where the persistence diagrams are encoded in $R$ and the generating cycles in the columns of $V$.
 To begin the reduction, one first  assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to $f$ into a $m \times m$  \emph{filtration boundary matrix} $D$. Setting $V = I$ and $R = D$, one proceeds by performing elementary left-to-right column operations on $V$ and $R$ until the following invariants are satisfied:
\\
\\
\textbf{ Decomposition Invariants:}
 \begin{enumerate}[labelsep=3pt, topsep=2pt,itemsep=-0.25ex,parsep=1.0ex]\label{inv:decomposition}
 	\item[I1.] $R = D V$ where $D$ is the boundary matrix of the filtration $(K,f)$
 	\item[I2.] $V$ is full-rank upper-triangular
 	\item[l3.] $R$ is \emph{reduced}: if $\mathrm{col}_i(R) \neq 0$ and $\mathrm{col}_j(R) \neq 0$, then $\mathrm{low}_R(i) \neq \mathrm{low}_R(j)$ 
 	\end{enumerate} 
 \vspace*{1em}
 where $\mathrm{low}_R(i)$ denotes the largest row index of a non-zero entry in column $i$ of $R$. 
We call the decomposition satisfying these three invariants \emph{valid}. The persistence diagrams of the corresponding filtration  can be determined from the lowest entries in $R$, once it has been reduced. Note that though the matrices $R$ and $V$ are not unique, the collection of persistent pairings are~\cite{zomorodian2005computing}.

@@

% the number of column operations has been observed to scale linearly in the size of the filtration in practice~\cite{bauer2021ripser}, due in part to the high sparsity and structure of $D$. 
Despite this high algorithmic complexity, many variations and optimizations to Algorithm~\ref{alg:reduce} have been proposed over the past decade, see~\cite{bauer2021ripser, bauer2017phat, chen2011persistent} for an overview. 

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, detecting homological critical events along the way.
As homological critical events only occur when the filtration order changes, detecting all such events may be reduced to computing valid decompositions at   time points interleaving all changes in the filtration order. 
For simplexwise filtrations, these changes manifest as transpositions of adjacent simplices, and thus any fixed set of rules that maintains a valid $R = D V$ decomposition under adjacent column transpositions is sufficient to compute persistence dynamically.

@@

 The time complexity of vineyards is determined entirely by the complexity of performing a single adjacent transposition.
Formally, since column operations are the largest complexity operations needed and each column can have potentially $O(m)$ entries, the complexity of vineyards is $O(m)$ per transposition. 
Inspection of the individual cases of the algorithm from~\cite{cohen2006vines} shows that any single transposition requires at most two such operations on both $R$ and $V$.
However, several factors can affect the runtime efficiency of the vineyards algorithm. 
On the positive side, as both the $V$ and $R$ matrices are often sparse, the cost of a given column operation is proportional to the number of non-zero  entries in the two columns being modified. Moreover, as a rule of thumb, its been observed that most transpositions require no column operations~\cite{edelsbrunner2000topological}. 
On the negative side, one needs to frequently query the non-zero status of various entries in $R$ and $V$ (consider evaluating e.g. Case 1.1 in~\cite{cohen2006vines}), which accrues a non-trivial runtime cost due to the quadratic frequency with which they are required.   
% Moreover, an efficient implementation of the vineyards algorithm requires a non-standard sparse matrix representation that allows swapping any two rows and columns in $O(1)$, as described in Figure 5 in~\cite{cohen2006vines}. 
% Though these minor operations do not affect the asymptotic complexity of the vineyards algorithm, they collectively accrue a non-trivial runtime cost

@@

For completeness, we recapitulate the motivation of the moves algorithm from~\cite{busaryev2010tracking}. 
Let $f: K \rightarrow [m]$ denote a filtration of size $m = \lvert K \rvert$ and $R = DV$ its decomposition, where $R = [r_1, r_2, \dots, r_m]$ and $V = [v_1, v_2, \dots, v_m]$ denote the columns of $R$ and $V$, respectively. By definition, if $r_j = D v_j = 0$, then the inclusion $K_{j-1} \hookrightarrow K_{j}$ introduced a new cycle whose generator is given by:
%the non-zero coefficients $(\alpha_i^{(j)}, \alpha_{i+1}^{(j)}, \dots, \alpha_j^{(j)})$ via the combination: 
\begin{equation}\label{eq:coefficients}
	r_j = D v_j = \alpha_j^{(1)} \cdot \partial(\sigma_{1}) + \alpha_{j}^{(2)} \cdot \partial(\sigma_{2}) + \dots + \alpha_{j}^{(j)} \cdot \partial(\sigma_{j}) = 0
\end{equation}
Now, consider the permutation $P$ that moves a simplex $\sigma_i$ in $K$ to position $j$, shifting all intermediate simplices $\sigma_{i+1}, \dots, \sigma_{j}$ down by one ($i < j$). 
To perform this shift, all of the non-zero coefficients $\alpha_k^{(i)}$ contributing to $Dv_k$ for $k \in [i+1, j]$ must be set to zero, as otherwise $\partial(\sigma_i)$ may participate in boundary chains earlier than its own birth in the permuted filtration. 
Setting these coefficients to zero amounts to setting row entries $v_k(i) = 0$ for all $k \in [i+1,j]$---these cancellations manifest as column operations $(\ast)$ used to ensure $P V(\ast)$ is upper-triangular, thus maintaining invariant I2.
Notice that these operations may yield an unreduced $R' = P R(\ast) P^T$, breaking invariant I3.
We could reduce $R'$ with an additional column operations, but the number of such operation scales $O(\lvert i - j\rvert^2)$, which no more efficient than simply performing the permutation and re-reducing the columns $[i,j]$ in $R$.

@@

\vspace{-2.2em}

@@

Perhaps surprisingly, small changes to set of allowable edit operations $\Sigma$ dramatically affect both the size of $d_\Sigma(p,q)$ and the difficulty of obtaining a minimal sorting. 
For example, while sorting by transpositions and reversals is NP-hard and sorting by prefix  transpositions is unknown, there are polynomial time algorithms for sorting by block interchanges, exchanges, and prefix exchanges~\cite{labarre2013lower}. Sorting by adjacent transpositions can be achieved in many ways: any sorting algorithm that exchanges two adjacent elements during its execution (e.g. bubble sort, insertion sort) yields a sorting of size $K_\tau(p, q)$.

@@

m_{ij} = \setlength\arraycolsep{1.8pt}

@@

m_{ij} = (\enspace i \enspace j \enspace j\text{-}1 \enspace \dots \enspace i\text{+}2 \enspace i\text{+}1 \enspace )

@@

However, in our setting, the total orders of any pair of simplexwise filtrations $(K, f), (K, f')$ of the same underlying complex $K$ may be thought of a permutations in $S_m$---the corresponding edit distance $d$ then reduces to the \emph{permutation edit distance} problem.

@@

Establishing a connection between the permutation edit distance and move scheduling allows us to exploit the combinatorial structure that comes from the developed theory on both LCS's and LIS's, which are both well-studied objects. We record a single corollary to demonstrate this fact. 

@@

\noindent Note the quantity from Corollary~\ref{cor:expectation} captures the size of $S^\ast$ between pairs of uniformly sampled permutations, as opposed to uniformly sampled filtrations, which have more structure due to the face poset. 

@@

The Spearman distance shares any similarities with $K_\tau$; like $K_\tau$, it is also metric on $S_m$ that is invariant under consistent relabeling. 

@@

%    \includegraphics[width=0.60\textwidth]{crossings.png} % arxiv only 
    \includegraphics[width=0.75\textwidth]{crossings.png} % springer only
